---
title: "Predicting_poverty"
author: "Farooq Qaiser"
date: "February 14, 2018"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# admin

```{r}

## set the seed to make rmd reproducible
set.seed(123)

```


## libraries 

```{r}

# don't want this mask margin from ggplot2
library(randomForest)

library(dplyr)
library(tidyr)
library(ggplot2)

library(dummies)

```


## load data

```{r}

data_loc = "/home/fqaiser94/poverty_data"

files = list.files(path = data_loc, pattern = '.*hhold.*train.*')	
file_paths = paste0(data_loc, '/', files)

for(file in file_paths) {
  
  print(file)
  
  raw_data = read.csv(
    file = file, 
    header = TRUE, 
    stringsAsFactors = FALSE
    )
  
}

raw_data

```

# EDA

## column names

```{r}

colnames(raw_data)

```

The majority of the columns' names have been masked. 
id, country, poor are the only ones I see which haven't been masked. 

## missing data

```{r}

ggplot_missing <- function(x){
  
  x %>% 
    is.na %>%
    reshape2::melt() %>%
    ggplot(
      aes(x = Var2, y = Var1)) +
    geom_raster(
      aes(fill = value)) +
    scale_fill_grey(
      name = "",
      labels = c("Present","Missing")) +
    theme_minimal() + 
    theme(
      axis.ticks = element_line(), 
      axis.text.x  = element_blank()) + 
    labs(
      x = "Variables",
      y = "Rows")
}

ggplot_missing(raw_data)

```

Absolutely no missing data. 

## class balance

```{r}

raw_data %>%
  ggplot(mapping = aes(x = poor)) +
  geom_bar() + 
  theme_minimal()

```

fairly balanced dataset

# Modelling

## preprocess data

```{r}

unique(raw_data$poor)

```

Our target variable appears to be encoded as strings. Let's double check that. 

```{r}

sapply(select(raw_data, poor), class)

```

yup, character.  

```{r}

raw_data$poor = ifelse(raw_data$poor=="True", TRUE, FALSE)
sapply(select(raw_data, poor), class)


```

changed to boolean

```{r}

which(raw_data=="True", arr.ind = TRUE)
which(raw_data=="False", arr.ind = TRUE)

```

double checked and verified that there are no other columns where True/False is encoded as strings

```{r}

unique(sapply(raw_data, class))

```

got four types of features in our data

```{r}

filter_by_column_type = function(df, type) {
  
  ind <- sapply(df, type)
  
  return(df[ind])
  
}

filter_by_column_type(head(raw_data), is.numeric) # integer or double (numeric)
filter_by_column_type(head(raw_data), is.character)


```

only 4 numeric (integer or double) columns (excluding id column)

```{r}

model_data = raw_data

```

create new df for preprocessed data

```{r}

# standardize numeric features
standardize <- function(df, omit_columns) {
 
  # identify columns for standardization
  ind = sapply(df, is.numeric)
  
  # omit specific columns for standardization
  ind[which(colnames(df) %in% omit_columns)] <- FALSE

  # standardize columns
  df[ind] = lapply(df[ind], scale)
  
  return(df)
}

model_data <- standardize(
  df = model_data, 
  omit_columns = 'id')


```

standardize numeric columns (excluding the id column)

```{r}

# OHE categorical features
model_data <- dummy.data.frame(
  data = model_data, 
  dummy.classes = c('character'),
  sep="_")

```

one hot encode character columns

## random forests model


```{r}

## 75% of the sample size
sample_size <- floor(0.75 * nrow(model_data))

train_ind <- sample(seq_len(nrow(model_data)), size = sample_size)

train <- model_data[train_ind, ]
test <- model_data[-train_ind, ]

```

split into train and test samples

```{r}

rf = randomForest(
  factor(poor) ~ . , 
  data = train 
  )

rf

```

```{r}

plot(rf)

```

```{r}

varImpPlot(rf)

```

## deep learning model

```{r}



```

